---
title: "Conditional transformations"
subtitle: "Practice with mutate() and case_when()"
author: ""
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: false
editor_options:
  chunk_output_type: console
---

## Task 1: Load packages

1.  Run the code chunk below to load the required packages.

```{r}
library(gapminder)
library(dplyr)
```

All exercises in this assignment use the `gapminder` data as a starting point.

2.  Run the code chunk that contains `glimpse(gapminder)`.

3.  How many variables are in the data set?

4.  How many observations are in the data set?

5.  What data types are contained in the variables?

```{r}
glimpse(gapminder)
```

## Task 2: A first use of `case_when()`

Run the following code chunk and observe the output in the Console. It adds a new column `gdp_group` to `gapminder` that classifies countries as `"low"`, `"middle"`, or `"high"` income based on GDP per capita.

```{r}
gapminder |>
  mutate(gdp_group = case_when(
    gdpPercap < 2000 ~ "low",
    gdpPercap < 15000 ~ "middle",
    gdpPercap >= 15000 ~ "high"
  ))
```

1.  Read the code out loud, verbally describing what each line does.

2.  Write in your own words what `case_when()` is doing here:

*Your answer here:*

## Task 3: Classify life expectancy

Use `mutate()` and `case_when()` to add a new column called `life_exp_group` to `gapminder` that classifies each row as:

- `"short"` when `lifeExp` is less than 50
- `"medium"` when `lifeExp` is between 50 and 70 (inclusive of 50, exclusive of 70)
- `"long"` when `lifeExp` is 70 or greater

```{r}
gapminder |>
  mutate(life_exp_group = case_when(
    lifeExp < 50 ~ "short",
    lifeExp >= 50 & lifeExp < 70 ~ "medium",
    lifeExp >= 70 ~ "long"
  ))
```

## Task 4: A default category with `TRUE`

Use `mutate()` and `case_when()` to add a new column called `pop_size` to `gapminder` that classifies each country as:

- `"small"` when `pop` is less than 1,000,000
- `"medium"` when `pop` is less than 50,000,000
- `"large"` for all remaining rows (use `TRUE ~ "large"` as the last condition)

```{r}
gapminder |>
  mutate(pop_size = case_when(
    pop < 1000000 ~ "small",
    pop < 50000000 ~ "medium",
    TRUE ~ "large"
  ))
```

1.  What does `TRUE ~ "large"` do? Why is it useful?

*Your answer here:*

## Task 5: Combining multiple conditions

Use `mutate()` and `case_when()` to add a new column called `status` to `gapminder` that classifies each row as:

- `"thriving"` when `lifeExp` is greater than 70 AND `gdpPercap` is greater than 15000
- `"developing"` when `lifeExp` is greater than 60 AND `gdpPercap` is greater than 2000
- `"struggling"` for everything else

Tip: use `&` to combine two conditions inside `case_when()`.

```{r}
gapminder |>
  mutate(status = case_when(
    lifeExp > 70 & gdpPercap > 15000 ~ "thriving",
    lifeExp > 60 & gdpPercap > 2000 ~ "developing",
    TRUE ~ "struggling"
  ))
```

## Task 6: Handling missing values

Run the following code chunk to create a small data frame with some missing values.

```{r}
practice_df <- tibble::tibble(
  country = c("A", "B", "C", "D", "E"),
  score = c(10, NA, 35, 70, NA)
)

practice_df
```

Use `mutate()` and `case_when()` to add a new column called `category` to `practice_df` that classifies each row as:

- `"low"` when `score` is less than 30
- `"high"` when `score` is 30 or greater
- `"unknown"` when `score` is `NA` (use `is.na(score) ~ "unknown"`)

```{r}
practice_df |>
  mutate(category = case_when(
    is.na(score) ~ "unknown",
    score < 30 ~ "low",
    score >= 30 ~ "high"
  ))
```

1.  What happens to `NA` values if you do *not* include `is.na(score) ~ "unknown"`?

*Your answer here:*

## Task 7: Save your work

In Tasks 3–6 the results appeared in the Console but did not get saved anywhere. Use the assignment operator `<-` to save your favourite version into a new object called `gapminder_classified`.

```{r}
gapminder_classified <- gapminder |>
  mutate(
    life_exp_group = case_when(
      lifeExp < 50 ~ "short",
      lifeExp >= 50 & lifeExp < 70 ~ "medium",
      lifeExp >= 70 ~ "long"
    ),
    pop_size = case_when(
      pop < 1000000 ~ "small",
      pop < 50000000 ~ "medium",
      TRUE ~ "large"
    )
  )

gapminder_classified
```

1.  What new object appears in your Environment?

2.  Why would we use the assignment operator `<-` to save our results?

## Task 8: Data communication

**In the YAML header (between the three dashes at the top of the document)**

1.  Add your name as the author of this document.

2.  Render the document and fix any errors.

## Task 9: Commit & Push & Issue

- Refer back to the instructions on the course website to complete the assignment.
