---
title: "Working with dates and times"
subtitle: "Practice with lubridate"
author: ""
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: false
editor_options:
  chunk_output_type: console
---

## Task 1: Load packages

1.  Run the code chunk below to load the required packages.

```{r}
library(lubridate)
library(dplyr)
library(ggplot2)
```

The `economics` dataset comes with `ggplot2` and contains US economic time series.

2.  Run the code chunk that contains `glimpse(economics)`.

3.  Which column already has a date type? What other columns are present?

```{r}
glimpse(economics)
```

## Task 2: Convert text to a date with `as_date()`

Run the following code chunk and observe the output. The function `as_date()` turns a character string into a proper `Date` object.

```{r}
as_date("2024-03-15")
```

1.  Use `as_date()` to convert each of the following strings into a date and run them in a single code chunk:

- `"2020-01-31"`
- `"2023-07-04"`
- `"1999-12-31"`

```{r}

```

2.  What class does `as_date()` return? Use `class()` to check on one of your results.

```{r}

```

## Task 3: What is today?

Run the code chunk below. The function `today()` returns the current date according to your computer's clock.

```{r}
today()
```

1.  Write code that calculates how many days have passed between `"2000-01-01"` and `today()`. Tip: you can subtract two dates with `-`.

```{r}

```

*Your answer (in days) here:*

## Task 4: Extract components with `year()`, `month()`, and `day()`

Use `mutate()` together with `year()`, `month()`, and `day()` to add three new columns to `economics`:

- `year` containing the year of `date`
- `month` containing the month of `date`
- `day` containing the day of the month of `date`

Save the result in a new object called `economics_dates`.

```{r}

```

1.  Look at the new columns. What data type are they?

*Your answer here:*

2.  Modify the `month()` call by adding `label = TRUE` to get month abbreviations (e.g. `"Jan"`, `"Feb"`) instead of numbers. Save the new result into `economics_dates` again.

```{r}

```

## Task 5: Day of the week with `wday()`

The function `wday()` returns the day of the week for a given date.

1.  Run the code chunk below and observe the output.

```{r}
wday("2024-03-15")
wday("2024-03-15", label = TRUE)
wday("2024-03-15", label = TRUE, week_start = 1)
```

2.  Add a new column called `weekday` to `economics_dates` (from Task 4) that contains the labelled day of the week of `date`, with the week starting on Monday (`week_start = 1`).

```{r}

```

## Task 6: Day of the year with `yday()`

The function `yday()` returns the day of the year (1 to 365 or 366).

1.  Run the code chunk below and predict the output before you run it.

```{r}
yday("2024-01-01")
yday("2024-12-31")
yday(today())
```

2.  Add a new column called `day_of_year` to `economics_dates` that contains the day of the year of `date`.

```{r}

```

3.  How many days into the year are we today?

*Your answer here:*

## Task 7: Filter by date components

Use `filter()` together with `year()` to create a subset of `economics` that only contains observations from the year 2010.

```{r}

```

Use `filter()` together with `year()` and `month()` to create a subset of `economics` that only contains observations from January of any year.

```{r}

```

## Task 8: Time zones with `with_tz()`

Dates with times (called *datetimes*) always belong to a time zone. The function `with_tz()` shows the same instant in a different time zone.

1.  Run the code chunk below and observe the output.

```{r}
moment <- ymd_hms("2024-06-21 12:00:00", tz = "UTC")
moment

with_tz(moment, tzone = "Europe/Zurich")
with_tz(moment, tzone = "America/New_York")
with_tz(moment, tzone = "Asia/Tokyo")
```

2.  Convert `moment` to the time zone `"Pacific/Auckland"`. What time and date is it there?

```{r}

```

*Your answer here:*

3.  `with_tz()` does *not* change the actual instant in time. Explain in your own words what it changes:

*Your answer here:*

## Task 9: Putting it together

Starting from `economics`, build a single pipeline that:

1.  Adds a column `year` with the year of `date`.

2.  Adds a column `month_name` with the labelled month of `date`.

3.  Filters to keep only observations from the year 2010 onwards.

4.  Saves the result to a new object called `economics_recent`.

```{r}

```

## Task 10: Data communication

**In the YAML header (between the three dashes at the top of the document)**

1.  Add your name as the author of this document.

2.  Render the document and fix any errors.

## Task 11: Commit & Push & Issue

- Refer back to the instructions on the course website to complete the assignment.
