---
title: "Communicating with tables"
subtitle: "Practice with gt() and kable()"
author: ""
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: false
editor_options:
  chunk_output_type: console
---

## Task 1: Load packages

1.  Run the code chunk below to load the required packages.

```{r}
library(gapminder)
library(dplyr)
library(knitr)
library(gt)
```

## Task 2: Create a small summary table

Run the code chunk below. It creates a small summary of life expectancy and GDP per capita by continent for the year 2007 and stores it in `gapminder_2007_summary`.

```{r}
gapminder_2007_summary <- gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    n_countries = n(),
    life_exp_mean = mean(lifeExp),
    gdp_mean = mean(gdpPercap)
  )

gapminder_2007_summary
```

1.  How many rows and columns does `gapminder_2007_summary` have?

*Your answer here:*

## Task 3: A first table with `kable()`

The function `kable()` from the `knitr` package prints a data frame as a nicely formatted table when the document is rendered.

1.  Run the code chunk below and then render the document to see the result.

```{r}
kable(gapminder_2007_summary)
```

2.  Compare the printed output in the Console with the rendered output. What is different?

*Your answer here:*

## Task 4: Control decimals with `digits`

The numbers in the table from Task 3 have many decimal places. Use the `digits` argument of `kable()` to round all numeric columns to 1 decimal place.

```{r}

```

## Task 5: Better column names

The column names `n_countries`, `life_exp_mean` and `gdp_mean` are fine for code but not ideal for a reader.

Use the `col.names` argument of `kable()` to display the columns as:

- `"Continent"`
- `"Number of countries"`
- `"Mean life expectancy (years)"`
- `"Mean GDP per capita (USD)"`

Keep `digits = 1` so values are still rounded to one decimal place.

```{r}

```

## Task 6: Add a caption with `kable()`

Use the `caption` argument of `kable()` to add the caption:

*"Life expectancy and GDP per capita by continent in 2007."*

Keep the nicer column names and the `digits = 1` argument from Task 5.

```{r}

```

1.  Render the document. Where does the caption appear in the output?

*Your answer here:*

## Task 7: A first table with `gt()`

The package `gt` provides a different way of building tables, with more options to customise the look.

1.  Run the code chunk below and render the document to see the result.

```{r}
gapminder_2007_summary |>
  gt()
```

2.  Briefly describe two differences between the `gt()` output and the `kable()` output:

*Your answer here:*

## Task 8: Add a title and a caption with `gt()`

Use `gt()` together with `tab_header()` to add:

- a title: `"Life expectancy and GDP per capita"`
- a subtitle: `"By continent, 2007"`

```{r}

```

## Task 9: Rename columns with `cols_label()`

Use `cols_label()` to give the columns reader-friendly names in your `gt()` table:

- `continent` → `"Continent"`
- `n_countries` → `"Number of countries"`
- `life_exp_mean` → `"Mean life expectancy (years)"`
- `gdp_mean` → `"Mean GDP per capita (USD)"`

Keep the title and subtitle from the previous task.

```{r}

```

## Task 10: Choose your favourite

Pick the table you like best from the previous tasks (either a `kable()` or a `gt()` version) and reproduce it in the code chunk below.

```{r}

```

1.  Why did you choose this version over the others? Think about how easy it is to read and how it would look in a report.

*Your answer here:*

## Task 11: Data communication

**In the YAML header (between the three dashes at the top of the document)**

1.  Add your name as the author of this document.

2.  Render the document and fix any errors.

## Task 12: Commit & Push & Issue

- Refer back to the instructions on the course website to complete the assignment.
