---
title: "Where Does the Time Go? A Time-Driven Activity-Based Costing Analysis of Informal Pit Latrine Emptying in Mzuzu, Malawi"
author: "Padraic Casserly"
date: today
format:
  html:
    toc: true
    toc-title: "Contents"
execute:
  echo: true
  warning: false
  message: false
bibliography: bibliography.bib
editor: visual
---

# Introduction

Informal pit latrine emptying is a critical but under-studied link in faecal sludge management (FSM) chains across Sub-Saharan African cities, where a majority of urban residents rely on on-site sanitation and manual emptiers to keep pits usable [@diop2017dakar; @brands2022nairobi]. Understanding the true cost of these informal services requires accurate data on how emptiers actually spend their time, since existing cost estimates are often built on interviews or rough assumptions rather than direct observation. Time-Driven Activity-Based Costing (TDABC) addresses this gap by decomposing a service into discrete tasks, timing each one directly, and multiplying observed durations by cost-driver rates, an approach that has previously been applied to sanitation-adjacent processes such as compost production in Malawi [@yesaya2021plastics; @kaplan2004tdabc]. This report applies the same logic to a TDABC study of informal pit latrine emptying ("gulper") services in Mzuzu, Malawi, conducted as part of doctoral research at ETH Zürich. Three ridealong observation sessions are combined into a single processed dataset here to characterise how gulpers' time is distributed across the tasks that make up an emptying job.

# Methods

Data were collected through structured ridealong observation, in which an enumerator timed one or more gulpers working in parallel on a single emptying job using a custom mobile data collection tool that logged the start and end of every interval. Here, an *interval* refers to one continuous row in the raw data: the length of time a single gulper spent on one subtask, from the moment they started it until they switched to a different subtask or task. Each of the three raw session files therefore records one row per interval: the gulper performing it, the high-level task and specific subtask, start and end clock times, duration in seconds, and a job identifier linking entries to their parent session. Because the collection tool pads each file with blank spacer rows and columns, the cleaning step below removes any column that is entirely missing and any row that is entirely missing before the three sessions are row-bound into a single combined dataset. Two columns are derived during cleaning, `job_date` (parsed from `job_id`) and `source_file`, to preserve session-level provenance, and the resulting analysis-ready dataset is written to `data/processed/processed_data.csv`.

## Task and subtask categories

The raw data organise every observed interval under one of seven high-level **tasks**, most of which are broken down into more specific **subtasks**:

-   **Earthworks** — physical work on the pit itself: digging a trench, filling the trench back in, cutting a hole in the pit liner, and repairing the liner afterward.
-   **Fluidizing** — preparing the sludge for removal by adding water: fetching water, adding it to the pit, and mixing it in.
-   **Extraction** — the core removal step: manually removing sludge from the pit and pulling out solid trash mixed in with it.
-   **Equipment** — handling tools and containers: maneuvering them into position, cleaning them, and occasional repairs.
-   **Superstructure** — work on the latrine structure above the pit: inspection, with occasional cleaning or repair.
-   **Communication** — coordination talk, either with the client or between gulpers on site.
-   **Rest** — recovery breaks between other tasks; recorded as a single undifferentiated category with no subtask breakdown.

## Install libraries

```{r}
#| label: setup

library(tidyverse)
library(knitr)
```

## Import and combine

```{r}
#| label: import-data

raw_dir <- "/cloud/project/data/raw"
raw_paths <- list.files(raw_dir, pattern = "\\.csv$", full.names = TRUE)

# Imports and cleans one raw session file: drops spacer columns/rows and
# tags the result with its source file and job date for provenance.
read_and_clean <- function(path) {
  read_csv(path, show_col_types = FALSE) %>%
    select(where(~ !all(is.na(.)))) %>%                # remove empty columns
    filter(if_any(everything(), ~ !is.na(.))) %>%       # remove empty rows
    mutate(
      source_file = basename(path),
      job_date = as.Date(str_extract(job_id, "^\\d{4}-\\d{2}-\\d{2}"))
    )
}

gulper_data <- map_dfr(raw_paths, read_and_clean)

glimpse(gulper_data)
```

## Save processed data

```{r}
#| label: save-processed

dir.create("/cloud/project/data/processed", showWarnings = FALSE)

write_csv(gulper_data, "/cloud/project/data/processed/processed_data.csv")
```

# Results

## Time by task category

```{r}
#| label: fig-task-time
#| fig-cap: "Total observed time by task category, summed across all three ridealong sessions."

task_time <- gulper_data %>%
  group_by(task) %>%
  summarise(total_minutes = sum(duration_seconds) / 60) %>%
  arrange(desc(total_minutes))

max_axis_minutes <- ceiling(max(task_time$total_minutes) / 30) * 30

ggplot(task_time, aes(x = fct_reorder(task, total_minutes), y = total_minutes, fill = task)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_viridis_d() +
  scale_y_continuous(breaks = seq(0, max_axis_minutes, by = 30)) +
  labs(
    x = "Task category",
    y = "Total observed time (minutes)",
    title = "Where gulpers' time goes"
  ) +
  theme_minimal()
```

@fig-task-time shows that Earthworks and Rest together account for roughly two-thirds of all observed time, while Communication, Fluidizing, Equipment and Superstructure make up the remainder despite occurring more often as short interludes between other tasks. This suggests that trench-digging and idle recovery time, rather than the sludge-handling tasks most associated with "emptying," are the largest time (and likely cost) drivers in a TDABC model of this service.

## Share of job time by task

```{r}
#| label: fig-task-share
#| fig-cap: "Percentage of each job's total observed time spent on each task category, for the three individual ridealong sessions (Job 1-3) plus a combined 'All jobs' column, separated from the individual jobs by a gap and a vertical divider line. Segment labels give the exact percentage for shares of 4% or more."

job_dates <- sort(unique(as.character(gulper_data$job_date)))
job_tick_labels <- paste0("Job ", seq_along(job_dates))

job_percent_data <- bind_rows(
  gulper_data %>% mutate(job_date = as.character(job_date)),
  gulper_data %>% mutate(job_date = "All jobs")
) %>%
  group_by(job_date, task) %>%
  summarise(total_seconds = sum(duration_seconds), .groups = "drop_last") %>%
  mutate(pct = 100 * total_seconds / sum(total_seconds)) %>%
  ungroup() %>%
  mutate(
    job_label = if_else(
      job_date == "All jobs",
      "All jobs",
      job_tick_labels[match(job_date, job_dates)]
    ),
    # a blank, data-free level between "Job 3" and "All jobs" creates a visual gap
    job_label = factor(job_label, levels = c(job_tick_labels, " ", "All jobs"))
  )

ggplot(job_percent_data, aes(x = job_label, y = pct, fill = task)) +
  geom_col() +
  geom_text(
    aes(label = if_else(pct >= 4, paste0(round(pct), "%"), "")),
    position = position_stack(vjust = 0.5),
    size = 3,
    colour = "white"
  ) +
  # dark vertical line marking the boundary between the individual jobs and "All jobs"
  geom_vline(xintercept = length(job_tick_labels) + 1, colour = "black", linewidth = 0.8) +
  scale_x_discrete(drop = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_fill_viridis_d() +
  labs(
    x = "Job",
    y = "Share of job time",
    fill = "Task category",
    title = "How a job's time breaks down, task by task"
  ) +
  theme_minimal()
```

@fig-task-share shows that Earthworks is the single largest share of every job (29-45%), but the balance of the rest of the time shifts a lot from job to job: Extraction and Fluidizing together used over 40% of Job 2 (31 May) but under 15% of Job 3 (13 June), while Rest was rare on Job 2 (3%) yet accounted for over 40% of Job 3. The combined "All jobs" column, set apart with its own gap, averages this out to roughly 35% Earthworks, 31% Rest, 11% Extraction, 9% Fluidizing, 8% Equipment, 5% Communication and 1% Superstructure, which is a useful overall benchmark but masks how much the mix actually swings from one job to the next. A single fixed cost-driver rate per task would therefore understate the cost of jobs with more digging and recovery time and overstate the cost of jobs dominated by fluidizing and extraction.

## Summary statistics

```{r}
#| label: tbl-task-summary
#| tbl-cap: "Summary statistics for task duration (seconds) by task category."

task_summary <- gulper_data %>%
  group_by(task) %>%
  summarise(
    n = n(),
    mean_sec = round(mean(duration_seconds), 1),
    median_sec = round(median(duration_seconds), 1),
    sd_sec = round(sd(duration_seconds), 1)
  ) %>%
  arrange(desc(n))

kable(
  task_summary,
  col.names = c("Task", "N intervals", "Mean (s)", "Median (s)", "SD (s)")
)
```

@tbl-task-summary confirms these patterns numerically: Earthworks has one of the highest mean durations among frequently observed tasks, while Communication is both frequent and brief. The standard deviations show that duration variability, and not just central tendency, differs markedly by task, which matters for setting realistic per-task cost-driver rates in a full TDABC costing model.

# Conclusions

-   Earthworks and Rest jointly account for roughly two-thirds of all observed gulper time across the three sessions, more than the combined Extraction and Fluidizing tasks conventionally seen as the "productive" core of pit emptying.
-   Task durations vary widely in both level and spread: Communication intervals are short and consistent, while Earthworks and Extraction are longer and more variable.
-   For TDABC-based costing of gulper services, earthworks and rest/recovery time deserve as much attention as the extraction and fluidizing tasks traditionally treated as the core of the job.
-   A natural next step would be to compare this task-duration data with how much money each job actually brings in (or pays the gulpers), to see whether jobs that take longer also earn more, and which specific tasks eat up time without adding much to a job's earnings.

# References

::: {#refs}
:::
