---
title: "index.qmd"
author: "Padraic Casserly"
format: html
editor: visual
---

## Project Description

The data collected for this project contain time duration and task information from a Time-Driven Activity-Based Costing (TDABC) study of informal pit latrine emptying services in Mzuzu, Malawi, conducted as part of doctoral research at ETH Zürich. TDABC is a costing methodology that estimates the operational cost of a service by decomposing it into the discrete tasks performed, measuring how long each task takes in practice, and then multiplying those observed durations by appropriate cost-driver rates (such as labour cost per unit time). Unlike traditional activity-based costing, which estimates time allocations through interviews or assumptions, TDABC relies on direct observation, making accurate task-duration data the central empirical input to the cost model.

The data was generated through structured ridealong observation in which an enumerator timed one or more "gulpers" (informal pit latrine emptiers) working in parallel on a single emptying job, using a custom-built mobile data collection tool that recorded the start and end of each task interval as gulpers moved through the job. Each row represents one continuous interval during which a single gulper was performing a single subtask: for example, a gulper might spend 92 seconds *fetching water* during the *Fluidizing* task (i.e., bringing water to the pit so that hardened sludge can be loosened), then 215 seconds *removing sludge* under *Extraction* (manually scooping sludge out of the pit), then 38 seconds in *Container transport* under *Sludge Handling* (moving the filled container to the truck). Higher-level tasks describe broad phases of the work — *Earthworks* (digging access trenches, cutting and later repairing the pit liner), *Equipment* (maneuvering, cleaning, and repairing tools and containers), *Communication* (with the client or between gulpers), *Travel* (to the disposal site), and *Rest* — while subtasks decompose those phases into the specific physical activities observed in the field.

The `gulper_id` and `entry_number` columns allow each interval to be attributed to a specific gulper and ordered within the session, and `job_id` links the entries to a parent session for which separately stored metadata (date, location, enumerator) is held privately.

## Install Libraries

```{r}
#| echo: true
#| message: false
#| warning: false

library(tidyverse)
```

## Import

```{r}
#| echo: true
#| message: false
#| warning: false

# Imports one of the datasets

may_21_data <- read_csv(
  "/cloud/project/data/raw/2026-05-21_01.csv",
  show_col_types = FALSE
)

glimpse(may_21_data)
```

## Clean

```{r}
#| echo: true

# Removes any spacer columns and rows filled with NA values

may_21_cleaned_data <- may_21_data %>%
  select(where(~ !all(is.na(.)))) %>%   # remove empty columns
  filter(if_any(everything(), ~ !is.na(.)))  # remove fully empty rows

glimpse(may_21_cleaned_data)
```
